---
title: "Assignment 6"
author: "Bernell Downer"
date: "11/16/2016"
output: html_document
---
#Addignment 6

In this assignment you will be looking at data from a MOOC. It contains the following per-student variables:

certified (yes/no) - Whether or not a student paid for the course  
forum.posts (numeric) - How many forum posts a student made throughout the course  
grade (numeric) - A student's average grade for the course exam  
assignment (numeric) - A student's average grade for the course assignments  

##Part I

#Packages
```{r}
#install.packages("rpart")
library(rpart)
```

#Data
```{r}
#Upload the data sets MOOC1.csv and MOOC2.csv
M1 <- read.csv("MOOC1.csv", header = TRUE)

M2 <- read.csv("MOOC2.csv", header = TRUE)

```

#Decision tree
```{r}
#Using the rpart package generate a classification tree predicting certified from the other variables in the M1 data frame. Which variables should you use?

# Grade and assignments variables are the best to use because students who pay for certification are more likel to be motivated and earn high average grades for the course exam and course assignments.

c.tree1 <- rpart(as.factor(certified) ~ grade + assignment,  method="class", data=M1)

#Check the results from the classifcation tree using the printcp() command

printcp(c.tree1)

#Plot your tree

post(c.tree1, file = "tree1.ps", title = "MOOC") #This creates a pdf image of the tree

```

##Part II

#The heading "xerror" in the printcp table stands for "cross validation error", it is the error rate of assigning students to certified/uncertified of the model averaged over 10-fold cross validation. CP stands for "Complexity Parameter" and represents the cost to error for adding a node to the tree. Notice it decreases as we add more nodes to the tree which implies that more nodes make better predictions. However, more nodes also mean that we may be making the model less generalizable, this is known as "overfitting".

#If we are worried about overfitting we can remove nodes form our tree using the prune() command, setting cp to the CP value from the table that corresponds to the number of nodes we want the tree to terminate at. Let's set it to two nodes.

```{r}
c.tree2 <- prune(c.tree1, cp = 0.058182)#Set cp to the level at which you want the tree to end

#Visualize this tree and compare it to the one you generated earlier

# c.tree2 has fewer branches than c.tree1. c.tree2 classifies 21 of 725 students as paying for certification, whereas c.tree1 classifies 5 of 725 students as not paying for certification.

post(c.tree2, file = "tree2.ps", title = "MOOC") #This creates a pdf image of the tree
```

#Now use both the original tree and the pruned tree to make predictions about the the students in the second data set. Which tree has a lower error rate?

```{r}
M2$predict1 <- predict(c.tree1, M2, type = "class")

M2$predict2 <- predict(c.tree2, M2, type = "class")

table(M2$certified, M2$predict1)
# Error rate = (24+7790)/10000 = 0.7814

table(M2$certified, M2$predict2)
# Error rate = (1184+3453)/10000 = 0.4637

```

##Part III

Choose a data file from the (University of Michigan Open Data Set)[https://github.com/bkoester/PLA/tree/master/data]. Choose an outcome variable that you would like to predict. Build two models that predict that outcome from the other variables. The first model should use raw variables, the second should feature select or feature extract variables from the data. Which model is better according to the cross validation metrics?

```{r}
D1 <- read.csv("student.course.csv", header = TRUE)

## 75% of the sample size
smp_size <- floor(0.75 * nrow(D1))

## set the seed to make your partition reproducible
set.seed(2191978)
train_ind <- sample(seq_len(nrow(D1)), size = smp_size)

# Training set
train <- D1[train_ind, ]

# Test set
test <- D1[-train_ind, ]

```

```{r}
# Assign 1 to overall GPA (GPAO) greater than 3.4 and assign 0 to overall GPA (GPAO) less than or equal to 3.4
train$GPAO[train$GPAO <= 3.4]=0
train$GPAO[train$GPAO > 3.4]=1

```

```{r}
#Using the rpart package generate a classification tree predicting GPAO honors eligibility (GPAO > 3.4) using three predictor variables — course grade (GRD_PTS_PER_UNIT), the term (TERM) the course was taught, and the academic division (DIV) to which the course belongs.

c.tree3 <- rpart(as.factor(GPAO) ~ GRD_PTS_PER_UNIT + TERM + DIV,  method="class", data=train)
```

```{r}
#Check the results from the classification tree using the printcp() command

printcp(c.tree3)

#Plot your tree

post(c.tree3, file = "tree3.ps", title = "MICHIGAN") #This creates a pdf image of the tree
```
```{r}
c.tree4 <- prune(c.tree3, cp = 0.01000)#Set cp to the level at which you want the tree to end

post(c.tree4, file = "tree4.ps", title = "MICHIGAN") #This creates a pdf image of the tree

```

```{r}
# Assign 1 to course grade (GRD_PTS_PER_UNIT) greater than 3.4 and assign 0 to course grade (GRD_PTS_PER_UNIT) less than or equal to 3.4
train$GRD_PTS_PER_UNIT[train$GRD_PTS_PER_UNIT <= 3.4]=0
train$GRD_PTS_PER_UNIT[train$GRD_PTS_PER_UNIT > 3.4]=1

```

```{r}
#Using the rpart package generate a classification tree predicting GPAO honors eligibility (GPAO > 3.4) using two predictor variables — course grade (GRD_PTS_PER_UNIT) and the term (TERM) the course was taught.

c.tree5 <- rpart(as.factor(GPAO) ~ GRD_PTS_PER_UNIT + TERM,  method="class", data=train)

```

```{r}
#Check the results from the classification tree using the printcp() command

printcp(c.tree5)

#Plot your tree

post(c.tree5, file = "tree5.ps", title = "MICHIGAN") #This creates a pdf image of the tree

```

```{r}
c.tree6 <- prune(c.tree5, cp = 0.01000)#Set cp to the level at which you want the tree to end

post(c.tree6, file = "tree6.ps", title = "MICHIGAN") #This creates a pdf image of the tree

```

```{r}
#Now use both the original tree and the pruned trees to make predictions about the the students in the test data set. Which tree has a lower error rate?

# Assign 1 to overall GPA (GPAO) greater than 3.4 and assign 0 to overall GPA (GPAO) less than or equal to 3.4
test$GPAO[test$GPAO <= 3.4]=0
test$GPAO[test$GPAO > 3.4]=1

test$predict3 <- predict(c.tree3, test, type = "class")
table(test$GPAO, test$predict3)
# Error rate = (42568+46810)/331767 = 0.2694

test$predict4 <- predict(c.tree4, test, type = "class")
table(test$GPAO, test$predict4)
# Error rate = (42568+46810)/331767 = 0.2694


# Assign 1 to course grade (GRD_PTS_PER_UNIT) greater than 3.4 and assign 0 to course grade (GRD_PTS_PER_UNIT) less than or equal to 3.4
test$GRD_PTS_PER_UNIT[test$GRD_PTS_PER_UNIT <= 3.4]=0
test$GRD_PTS_PER_UNIT[test$GRD_PTS_PER_UNIT > 3.4]=1

test$predict5 <- predict(c.tree5, test, type = "class")
table(test$GPAO, test$predict5)
# Error rate = (41253+49050)/331767 = 0.2723

test$predict6 <- predict(c.tree6, test, type = "class")
table(test$GPAO, test$predict6)
# Error rate = (41253+49050)/331767 = 0.2723


# CONCLUSION: The original classification tree c.tree3 predicting GPAO honors eligibility (GPAO > 3.4) using three predictor variables — course grade (GRD_PTS_PER_UNIT), the term (TERM) the course was taught, and the academic division (DIV) to which the course belongs — has the lowest error rate = 0.2694

```


### To Submit Your Assignment

Please submit your assignment by first "knitting" your RMarkdown document into an html file and then commit, push and pull request both the RMarkdown file and the html file.